A Logical Solution of the Linearity Challenge

### 2025-11-27 paper/Type.lagda.tex
prose for types

In fact, this property is so important in the rest of the formalization that it
is worth defining an implicit rewriting rules so that Agda can autonomously
apply it whenever possible. This is achieved by means of the following pragma
directive.

@@

>>>>

In fact, this property is so important in the rest of the formalization that it
is worth defining an implicit rewriting rule so that Agda can autonomously apply
it whenever possible. This is achieved by means of the following pragma
directive.

@@

\begin{code}[hide]
ext : ∀{m n} → (Fin m → Fin n) → Fin (suc m) → Fin (suc n)
ext ρ zero = zero
ext ρ (suc k) = suc (ρ k)

rename : ∀{m n} → (Fin m → Fin n) → PreType m → PreType n
rename ρ 𝟘 = 𝟘
rename ρ 𝟙 = 𝟙
rename ρ ⊥ = ⊥
rename ρ ⊤ = ⊤
rename ρ (var x) = var (ρ x)
rename ρ (rav x) = rav (ρ x)
rename ρ (¡ A) = ¡ (rename ρ A)
rename ρ (¿ A) = ¿ (rename ρ A)
rename ρ (A & B) = rename ρ A & rename ρ B
rename ρ (A ⊕ B) = rename ρ A ⊕ rename ρ B
rename ρ (A ⊗ B) = rename ρ A ⊗ rename ρ B
rename ρ (A ⅋ B) = rename ρ A ⅋ rename ρ B
rename ρ ($∀ A) = $∀ (rename (ext ρ) A)
rename ρ ($∃ A) = $∃ (rename (ext ρ) A)

exts : ∀{m n} → (Fin m → PreType n) → Fin (suc m) → PreType (suc n)
exts σ zero = var zero
exts σ (suc k) = rename suc (σ k)
\end{code}

Next we define the function \AgdaFunction{subst} that simultaneously substitutes
the type variables of a (pre)type with other (pre)types. Even though in practice
we will always substitute one variable at a time, it is technically easier to
substitutes \emph{all} variables of a (pre)type. The definition of
\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
variables and \emph{lifting} substitutions across quantifiers. All such
definitions are straightforward adaptations of those describe by
\cite{KokkeWadler}.

\begin{code}
subst : ∀{m n} → (Fin m → PreType n) → PreType m → PreType n
\end{code}
\begin{code}[hide]
subst σ 𝟘 = 𝟘
subst σ 𝟙 = 𝟙
subst σ ⊥ = ⊥
subst σ ⊤ = ⊤
subst σ (var x) = σ x
subst σ (rav x) = dual (σ x)
subst σ (¡ A) = ¡ (subst σ A)
subst σ (¿ A) = ¿ (subst σ A)
subst σ (A & B) = subst σ A & subst σ B
subst σ (A ⊕ B) = subst σ A ⊕ subst σ B
subst σ (A ⊗ B) = subst σ A ⊗ subst σ B
subst σ (A ⅋ B) = subst σ A ⅋ subst σ B
subst σ ($∀ A) = $∀ (subst (exts σ) A)
subst σ ($∃ A) = $∃ (subst (exts σ) A)
\end{code}

Among all substitutions, we will use the one that substitutes the 0-indexed type
variable with a (pre)type. It is convenient to introduce this substitution once
and for all, which we do here.

\begin{code}
make-subst : ∀{n} → PreType n → Fin (suc n) → PreType n
make-subst A zero = A
make-subst A (suc k) = var k
\end{code}

Duality and substitutions are meant to commute.

\begin{code}
dual-subst : ∀{m n} {σ : Fin m → PreType n} {A : PreType m} →
             subst σ (dual A) ≡ dual (subst σ A)
\end{code}
\begin{code}[hide]
dual-subst {_} {_} {σ} {𝟘} = refl
dual-subst {_} {_} {σ} {𝟙} = refl
dual-subst {_} {_} {σ} {⊥} = refl
dual-subst {_} {_} {σ} {⊤} = refl
dual-subst {_} {_} {σ} {var x} = refl
dual-subst {_} {_} {σ} {rav x} = refl
dual-subst {_} {_} {σ} {¡ A} = cong ¿ (dual-subst {σ = σ} {A})
dual-subst {_} {_} {σ} {¿ A} = cong ¡ (dual-subst {σ = σ} {A})
dual-subst {_} {_} {σ} {A & B} = cong₂ _⊕_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
dual-subst {_} {_} {σ} {A ⊕ B} = cong₂ _&_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
dual-subst {_} {_} {σ} {A ⊗ B} = cong₂ _⅋_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
dual-subst {_} {_} {σ} {A ⅋ B} = cong₂ _⊗_ (dual-subst {σ = σ} {A}) (dual-subst {σ = σ} {B})
dual-subst {_} {_} {σ} {$∀ A} = cong $∃ (dual-subst {σ = exts σ} {A})
dual-subst {_} {_} {σ} {$∃ A} = cong $∀ (dual-subst {σ = exts σ} {A})
\end{code}

Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
formalization that follows. Therefore, we add it to the set of implicit
rewriting rules used by Agda so that we do not have to think about it again.

\begin{code}
{-# REWRITE dual-subst #-}
\end{code}

At this stage we are finally ready to define the type of closed pre-types, those
having no free type variables. We call \AgdaDatatype{Type} closed pre-types.
Hereafter, we will seldom use pre-types again.

\begin{code}
Type : Set
Type = PreType zero
\end{code}
